---
title: "Data Cleaning"
author: "Sahana Rayan"
date: "2023-09-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Install libraries if need be

```{r}
# install.packages("readxl")
```


# Load Libraries

```{r}
library(tidyverse)
library(readxl)
```

 
# Data Loading

```{r}
# Path for policy data
policy_file_path <- "data_policy_covid_economic_raw/OxCGRT_compact_subnational_v1.csv"
# Path for Unemployment data for Michigan
michigan_unemployment <- "data_policy_covid_economic_raw/unemployment/Michigan_NS_Unemployment.xlsx"
utah_unemployment <- "data_policy_covid_economic_raw/unemployment/Utah_NS_Unemployment.xlsx"
cali_unemployment <- "data_policy_covid_economic_raw/unemployment/California_NS_Unemployment.xlsx"
washington_unemployment <- "data_policy_covid_economic_raw/unemployment/Washington_NS_Unemployment.xlsx"
oklahoma_unemployment <- "data_policy_covid_economic_raw/unemployment/Oklahoma_NS_Unemployment.xlsx"

covid_cases_file_path <- "data_policy_covid_economic_raw/covid_data/covid_confirmed_usafacts.csv"
covid_deaths_file_path <- "data_policy_covid_economic_raw/covid_data/covid_deaths_usafacts.csv"
covid_population_file_path <- "data_policy_covid_economic_raw/covid_data/covid_county_population_usafacts.csv"
```


## Policy data
```{r}
policy_world_data <- read.csv(policy_file_path)
policy_us_data <- policy_world_data %>% 
  filter(CountryCode == "USA")
rm(policy_world_data)
```


## Function - Filter Data based on State of interest

```{r}
filter_by_state_policy <- function(state){
  policy_state_data <- policy_us_data %>% 
    filter(RegionCode == state)
  return(policy_state_data)
}
```

## Actual filtering on state(s)

```{r}
policy_mi_data <- filter_by_state_policy("US_MI")
policy_ok_data <- filter_by_state_policy("US_OK")
policy_ca_data <- filter_by_state_policy("US_CA")
policy_ut_data <- filter_by_state_policy("US_UT")
policy_wa_data <- filter_by_state_policy("US_WA")
```



## Unemployment data for Michigan

```{r}
unemployment_mi_data <- read_excel(michigan_unemployment)
unemployment_ok_data <- read_excel(oklahoma_unemployment)
unemployment_ca_data <- read_excel(cali_unemployment)
unemployment_ut_data <- read_excel(utah_unemployment)
unemployment_wa_data <- read_excel(washington_unemployment)
```

## Covid data 

```{r}
covid_cases <- read.csv(covid_cases_file_path)
covid_population <- read.csv(covid_population_file_path)
covid_deaths <- read.csv(covid_deaths_file_path)
```

## Function - Filter Data based on State of interest

```{r}
filter_by_state_covid <- function(data_covid, state){
  covid_state_data <- data_covid %>% 
    filter(State == state)
  return(covid_state_data)
}
```

```{r}
covid_mi_cases <- filter_by_state_covid(covid_cases, "MI")
covid_mi_population <- filter_by_state_covid(covid_population, "MI")
covid_mi_deaths <- filter_by_state_covid(covid_deaths, "MI")

covid_ok_cases <- filter_by_state_covid(covid_cases, "OK")
covid_ok_population <- filter_by_state_covid(covid_population, "OK")
covid_ok_deaths <- filter_by_state_covid(covid_deaths, "OK")

covid_ca_cases <- filter_by_state_covid(covid_cases, "CA")
covid_ca_population <- filter_by_state_covid(covid_population, "CA")
covid_ca_deaths <- filter_by_state_covid(covid_deaths, "CA")

covid_ut_cases <- filter_by_state_covid(covid_cases, "UT")
covid_ut_population <- filter_by_state_covid(covid_population, "UT")
covid_ut_deaths <- filter_by_state_covid(covid_deaths, "UT")

covid_wa_cases <- filter_by_state_covid(covid_cases, "WA")
covid_wa_population <- filter_by_state_covid(covid_population, "WA")
covid_wa_deaths <- filter_by_state_covid(covid_deaths, "WA")
```




# Subsetting and Cleaning of Policy Data

C2M_Workplace.closing is an ordinal variable which takes on these values
0 - no measures
1 - recommend closing (or recommend work from home) or all businesses open with alterations resulting in significant differences compared to non-Covid-19 operation
2 - require closing (or work from home) for some sectors or categories of workers
3 - require closing (or work from home) for all-but-essential workplaces (eg grocery stores, doctors)
Blank - no data

C2M_flag is an indicator variable which means
0 - targeted (for the state specifically)
1- general (for the entire country)
Blank - no data

## Function to do the cleaning

```{r}
clean_policy_data <- function(dataset_policy) {
  policy_data_clean <- dataset_policy %>% 
    select(c(Date, C2M_Workplace.closing, C2M_Flag)) %>% 
    mutate(Date = ymd(Date)) %>% 
    mutate(Year = year(Date), Month = month(Date)) %>% 
    group_by(Year, Month) %>% 
    summarise(num_days = n(),
              workplace_0 = sum(C2M_Workplace.closing == 0),
              workplace_1 = sum(C2M_Workplace.closing == 1),
              workplace_2 = sum(C2M_Workplace.closing == 2),
              workplace_3 = sum(C2M_Workplace.closing == 3),
              workplace_blank = sum(is.na(C2M_Workplace.closing)),
              flag_0 = sum(C2M_Flag == 0),
              flag_1 = sum(C2M_Flag == 1),
              flag_blank = sum(is.na(C2M_Flag))) %>% 
    mutate(flag_0 = ifelse(is.na(flag_0), 0, flag_0),
           flag_1 = ifelse(is.na(flag_1), 0, flag_1))
  return(policy_data_clean)
}
```

 
## Run it on a specific state(s)

```{r}
policy_mi_data_clean <- clean_policy_data(policy_mi_data)
policy_ok_data_clean <- clean_policy_data(policy_ok_data)
policy_ca_data_clean <- clean_policy_data(policy_ca_data)
policy_ut_data_clean <- clean_policy_data(policy_ut_data)
policy_wa_data_clean <- clean_policy_data(policy_wa_data)

```


# Subsetting and Cleaning Unemployment data

## Function to do it

```{r}
clean_unemployment_data <- function(dataset_unemployment) {
  unemployment_data_clean <- dataset_unemployment %>% 
    filter(Year != 2019) %>% 
    rename("Month" = "Period") %>% 
    mutate(Month = recode(Month,
                          Jan = 1, Feb = 2,Mar = 3, Apr = 4, May = 5, Jun = 6,
                          Jul = 7, Aug = 8, Sep = 9, Oct = 10, Nov = 11, Dec = 12))
}
```

## Run it on specific states

```{r}
unemployment_mi_clean <- clean_unemployment_data(unemployment_mi_data)
unemployment_ok_clean <- clean_unemployment_data(unemployment_ok_data)
unemployment_ca_clean <- clean_unemployment_data(unemployment_ca_data)
unemployment_ut_clean <- clean_unemployment_data(unemployment_ut_data)
unemployment_wa_clean <- clean_unemployment_data(unemployment_wa_data)

```

# Merging the two data frames

## Function to do it

```{r}
merge_data <- function(unemployment, policy) {
  merged <- inner_join(unemployment, policy,
                       by = join_by("Year" == "Year", "Month" == "Month") )
}
```


## Actually doing it on the dataframe

```{r}
merged_mi_data <- merge_data(unemployment_mi_clean, policy_mi_data_clean)
merged_ok_data <- merge_data(unemployment_ok_clean, policy_ok_data_clean)
merged_ca_data <- merge_data(unemployment_ca_clean, policy_ca_data_clean)
merged_ut_data <- merge_data(unemployment_ut_clean, policy_ut_data_clean)
merged_wa_data <- merge_data(unemployment_wa_clean, policy_wa_data_clean)

```


# Subsetting, merging, and cleaning covid data


```{r}
clean_covid_data <- function(covid_data) {
  clean_covid_data <- covid_data %>% 
    pivot_longer(cols = 5:ncol(covid_data),
                 names_to = "Date",
                 values_to = "Value") %>% 
    group_by(Date) %>% 
    summarise(sum_value = sum(Value)) %>% 
    mutate(Date = ymd(str_sub(Date, start = 2, end = -1)))  %>% 
    mutate(new_value = sum_value - lag(sum_value, default = 0)) %>%
    mutate(Year = year(Date), Month = month(Date)) %>%
    group_by(Year, Month) %>%
    summarise(sum_val = sum(new_value))
  return(clean_covid_data)
}
```

```{r}
covid_mi_cases_clean <- clean_covid_data(covid_mi_cases) %>% 
  rename("Case Count" = sum_val)
covid_mi_deaths_clean <- clean_covid_data(covid_mi_deaths) %>% 
  rename("Covid Deaths" = sum_val)
covid_mi_clean <- inner_join(covid_mi_cases_clean, covid_mi_deaths_clean, 
                             by = join_by("Year" == "Year", "Month" == "Month"))

covid_ok_cases_clean <- clean_covid_data(covid_ok_cases) %>% 
  rename("Case Count" = sum_val)
covid_ok_deaths_clean <- clean_covid_data(covid_ok_deaths) %>% 
  rename("Covid Deaths" = sum_val)
covid_ok_clean <- inner_join(covid_ok_cases_clean, covid_ok_deaths_clean, 
                             by = join_by("Year" == "Year", "Month" == "Month"))

covid_ca_cases_clean <- clean_covid_data(covid_ca_cases) %>% 
  rename("Case Count" = sum_val)
covid_ca_deaths_clean <- clean_covid_data(covid_ca_deaths) %>% 
  rename("Covid Deaths" = sum_val)
covid_ca_clean <- inner_join(covid_ca_cases_clean, covid_ca_deaths_clean, 
                             by = join_by("Year" == "Year", "Month" == "Month"))

covid_ut_cases_clean <- clean_covid_data(covid_ut_cases) %>% 
  rename("Case Count" = sum_val)
covid_ut_deaths_clean <- clean_covid_data(covid_ut_deaths) %>% 
  rename("Covid Deaths" = sum_val)
covid_ut_clean <- inner_join(covid_ut_cases_clean, covid_ut_deaths_clean, 
                             by = join_by("Year" == "Year", "Month" == "Month"))

covid_wa_cases_clean <- clean_covid_data(covid_wa_cases) %>% 
  rename("Case Count" = sum_val)
covid_wa_deaths_clean <- clean_covid_data(covid_wa_deaths) %>% 
  rename("Covid Deaths" = sum_val)
covid_wa_clean <- inner_join(covid_wa_cases_clean, covid_wa_deaths_clean, 
                             by = join_by("Year" == "Year", "Month" == "Month"))
```


# Final merge

```{r}
merged_mi_data <- merge_data(merged_mi_data, covid_mi_clean)
merged_ok_data <- merge_data(merged_ok_data, covid_ok_clean)
merged_ca_data <- merge_data(merged_ca_data, covid_ca_clean)
merged_ut_data <- merge_data(merged_ut_data, covid_ut_clean)
merged_wa_data <- merge_data(merged_wa_data, covid_wa_clean)

```


# Save data

```{r}
directory <- "data_clean/"
write.csv(merged_mi_data, file = paste0(directory, "Michigan_cleaned.csv"))
write.csv(merged_ok_data, file = paste0(directory, "Oklahoma_cleaned.csv"))
write.csv(merged_ca_data, file = paste0(directory, "California_cleaned.csv"))
write.csv(merged_ut_data, file = paste0(directory, "Utah_cleaned.csv"))
write.csv(merged_wa_data, file = paste0(directory, "Washington_cleaned.csv"))

```














